%!TEX encoding = UTF-8 Unicode

%% Thesis cover pages
%%

% Remove 'coverdepname' parameter to remove department name from cover pages
% Add 'final' parameter to remove "DOCUMENTO PROVISÓRIO" from cover page
\documentclass[11pt,twoside,a4paper,final]{report}

%%% This file is a template for the cover of your
%%%  thesis.
%%% Lines like this one, starting with %%% , are
%%%  instructions. Search them to know where and
%%%  how to edit this file. After editing the
%%%  document you can delete them.
%%% Items marked with OPTIONAL can be removed.
%%% Most of the lines you will have to change
%%%  already have examples to help you understand.
%%% The complete list of options and commands can
%%%  be found in uaThesisTemplate.sty


%%% Select your area of study and department from
%%%  the following lists and replace them in the
%%%  \usepackage declaration ([area,department]).
%%%
%%% | area        |   | department |
%%% ---------------   --------------
%%% | arts        |   | dao        |
%%% | sciences    |   | dbio       |
%%% | education   |   | dcspt      |
%%% | economy     |   | deca       |
%%% | engineering |   | degeit     |
%%% | humanities  |   | de         |
%%% | health      |   | deti       |
%%% |             |   | demac      |
%%% |             |   | decivil    |
%%% |             |   | dem        |
%%% |             |   | fis        |
%%% |             |   | dlc        |
%%% |             |   | dmat       |
%%% |             |   | dq         |
%%% |             |   | dcm        |
%%% |             |   | geo        |
%%% ---------------   --------------
%% Import the UA thesis template
\usepackage[engineering,deti]{uaThesisTemplate}


%% Optional packages for utf8 encoding and T1 fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%% For prettier pages
\usepackage{lmodern}

%% We will use english and portuguese
\usepackage[english,portuguese]{babel}

\usepackage{xspace}% used by \sigla


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Edit thesis year
%% Thesis Year
\ThesisYear{\the\year}


%%% Uncomment one of the following
%\DocumentTypeThesis
\DocumentTypeDissertation

%%% Edit author's name
%% Author's full name
\Author{Mário Francisco\\Costa Silva}

%%% Edit thesis title in Portuguese
%% Thesis Mandatory Title in Portuguese
\Title{\selectlanguage{portuguese}Avaliação de Áudio de Dados Afetivos em Sessões de Videoconferência}


%%% OPTIONAL: you can write your thesis title in
%%%  other languages. You can repeat the following
%%%  line as many times as you want, or remove it
%%%  if you only want the portuguese title.
%% Optional Titles in additional languages.
\Title{\selectlanguage{english}Audio Evaluation of Affective Data in Videoconference Sessions}

%%% OPTIONAL: add a quotation if you like.
%% Optional Quotation
\Quote{\selectlanguage{english}Just like we can understand speech and machines can communicate in speech, we also understand and communicate with humor and other kinds of emotions. And machines that can speak the language of emotions are going to have better, more effective interactions with us}{MIT Sloan professor Erik Brynjolfsson}

%%% Edit your supervisor's information.
%%%
%%% \Supervisor{name}{category}
%%%
%%% name   : Supervisor's name
%%% categoy: One of the following list
%%%
%%% | category               |
%%% --------------------------
%%% | catedrático            |
%%% | catedrático convidado  |
%%% | associado c/ agregação |
%%% | associado              |
%%% | auxiliar               |
%%% | auxiliar convidado     |
%%% --------------------------
%% Supervisor information
\Supervisor{Susana Manuela Martinho dos Santos Baía Brás}{Investigadora no Instituto de Engenharia Eletrónica e Telemática de Aveiro}

%%% OPTIONAL: you can add a second supervisor.
%%%
%%% \SecondSupervisor{name}{category}
%%%  {[preposition: (at, in, no, da)] department}
%%%  {[preposition: (at, in, do, na)] school}
%%%
%%% name      : See above.
%%% category  : See above.
%%% department: Can be a department, faculty or
%%%             other; Choose the preposition
%%%             accordingly.
%%% school    : University/Scholl name; Choose the
%%%             preposition accordingly.
%% Optional SecondSuperviser information
\SecondSupervisor{Ilídio Castro Oliveira}{Auxiliar}{do Departamento de Eletrónica, Telecomunicações e Informática}

%%% Edit your degree information.
%%%
%%% \Degree{title}{name}
%%%
%%% title: Degree title (Mestre or Doutor)
%%% name : Degree name
%% Degree title and name
\Degree{Mestre}{Engenharia Informática}

%%% OPTIONAL: edit grant support texts where
%%%  appropriate. Maximum two.
%%%
%%% \GrantText{grant1}{grant2}
%% Optional Grant support texts
%\GrantText{Texto Apoio financeiro do POCTI no âmbito do III Quadro Comunitário de Apoio.}
          %{Texto Apoio financeiro da FCT e do FSE no âmbito do III Quadro Comunitário de Apoio.}

%%% OPTIONAL: Add a dedication text.
%% Optional Dedication text
%\Dedication{Dedico este trabalho à minha esposa e filho pelo incansável apoio.}

%%% OPTIONAL: Add an acknowledgement text
%% Optional Acknowledgement text
%\Acknowledgements{Agradeço toda a ajuda a todos os meus colegas e companheiros.}

%%% Edit the keywords and abstract in Portuguese
%%% Don't forget to remove the \blindtext[3]
%%%  declaration.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation
%% Mandatory Keywords and Abstract in Portuguese
\Abstract{Palavras Chave}{Computação Afetiva, Processamento de Voz, Reconhecimento de Emoção da Fala, Aprendizagem Automática}
         {Resumo}{Esta disseratação apresenta um estudo abrangente do Reconhecimento de Emoção de Fala (REF) usando abordagens tradicionais de aprendizagem automática e aprendizagem profunda. A principal contribuição deste trabalho é o desenvolvimento e avaliação de dois modelos em múltiplos datasets. Além de explorar o impacto de diferentes conjuntos de características do áudio e metodologias de validação, também investigamos a importância das técnicas de pré-processamento de áudio e seu efeito no desempenho do modelo. Por meio de estudos experimentais, desenvolvemos dois modelos para REF: um modelo eXtreme Gradient Boosting (XGBoost) para a abordagem tradicional utilizando um vetor unidimensional de 33 características do áudio e um modelo ResNet50 ajustado usando imagens de espectrograma para aprendizagem profunda. Estes modelos alcançaram precisões de 60,69\% e 58,24\%, respectivamente, para validação cruzada estratificada de 5 vezes no dataset Interactive Emotional Dyadic Motion Capture (IEMOCAP). Além disso, o modelo de aprendizagem profunda superou o modelo tradicional na avaliação cruzada de datasets devido à sua maior capacidade de extração e generalização de recursos, enquanto o modelo tradicional é mais adequado para aplicações em tempo real devido à sua velocidade de processamento mais rápida. Além disso, uma análise detalhada dos dados usando várias estratégias de estratificação levou à identificação de um conjunto de condições para o IEMOCAP que melhorou o desempenho geral dos modelos na avaliação cruzada de datasets. Também foi desenvolvido um pipeline que automatiza o processo de REF para classificação em tempo real ou offline, criando segmentos de áudio com uma determinada duração e classificando as emoções presentes neles usando os modelos desenvolvidos. No geral, esta disseratação fornece uma base sólida para pesquisas futuras no desenvolvimento de modelos REF mais robustos e precisos, oferecendo uma implementação abrangente e processo de pensamento, juntamente com conclusões e interpretações dos resultados obtidos. As nossas conclusões contribuem para o crescente corpo de pesquisa de REF e fornecem informações valiosas para investigadores e profissionais da área.}

%%% OPTIONAL: you can write the keywords and
%%%  abstract in other languages. You can repeat
%%%  the following lines as many times as you want
%%%  or remove them if you only want portuguese.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation.
%%% You can modify the language to any other supported by Latex
%% Optional Keywords and Abstract in additional languages.
\Abstract{Keywords}{Affective Computing, Voice Processing, Speech Emotion Recognition, Machine Learning}
         {Abstract}{This dissertation presents a comprehensive study of Speech Emotion Recognition (SER) using traditional machine learning and deep learning approaches. The main contribution of this work is the development and evaluation of two models on multiple datasets. In addition to exploring the impact of different feature sets and validation methodologies, we also investigate the importance of audio preprocessing techniques and their effect on model performance. Through experimental studies, we developed two models for SER: an eXtreme Gradient Boosting (XGBoost) model for the traditional approach utilizing a 1-dimensional vector of 33 audio features, and a fine-tuned ResNet50 model using spectrogram images for deep learning. These models achieved accuracies of 60.69\% and 58.24\%, respectively, for 5-fold stratified cross-validation on the Interactive Emotional Dyadic Motion Capture (IEMOCAP) dataset. Moreover, the deep learning model outperformed the traditional model in the cross-dataset evaluation due to its higher feature extraction and generalization capabilities, while the traditional model is more suitable for real-time applications due to its faster processing speed. Furthermore, a detailed analysis of the data using several stratification strategies led to the identification of a set of conditions for IEMOCAP that improved the general performance of the models in cross-dataset evaluation. A pipeline was also developed that automates the SER process for both real-time and offline classification by creating voiced audio segments with a certain duration and classifying the emotions present in them using the developed models. Overall, this dissertation provides a solid foundation for future research in developing more robust and accurate SER models, by offering a comprehensive implementation and thought process, along with conclusions and interpretations of the obtained results. These findings contribute to the growing body of research on SER and provide valuable insights for researchers and practitioners in the field.}

%%% Edit the Jury President information.
%% First Member will be the President of the Jury
\JuriMember{Prof. Doutor João Antunes da Silva}{professor associado da Faculdade de Engenharia da Universidade do Porto}

%%% Repeat the above declaration for additional
%%%  members of the jury as many times as necessary.
%% Additional members of the Jury
\JuriMember{Prof. Doutor João Antunes da Silva}{professor associado da Faculdade de Engenharia da Universidade do Porto}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  END
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%%% By default we consider portuguese to be the
%%%  main language (the last one in the usepackage
%%%  declaration). If you're writing your thesis
%%%  in english uncomment the following line.
%%% Set the proper language or you will end with incorrect hyphenation!
\selectlanguage{english}

\MakeCover

\end{document}
